---
title: "CDS_BDSI_Project_GGKADY"
date: "2024-07-10"
output: html_document
---
BDSI Cancer Data Science Project


Data Cleaning & Preparation
```{r setup, include=FALSE}
library(survival)
library(ranger)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(survminer)
library(glmnet)
library(tidyr)
library(rjags)
library(afthd)
library(penAFT)

#Load lung meta data
lungMeta<-read.csv("newLungKData.csv")
lungMeta

#filter image with most cells per patient
lungMeta
modData <- lungMeta %>%
  group_by(patient_id) %>%
  slice(which.max(totalCell)) %>%
  ungroup()

#recode demographic variables
#mhcII <- mchII_status
# 0-low, 1-high

#gender_1 <- gender
# 0- male, 1- female


modData1 <- modData %>%
  mutate(mchII= case_when(mhcII_status=="low" ~ 0,
                          mhcII_status=="high" ~ 1),
         gender_1= case_when(gender=="F" ~ 0,
                             gender=="M" ~ 1),
         age_at_diagnosis = as.numeric(age_at_diagnosis), 
         stage_new = case_when(stage_numeric==1 ~ "1",
                               stage_numeric==2 ~ "2",
                               stage_numeric==3 ~ "3_plus",
                               stage_numeric==4 ~ "3_plus"))
modData1$stage_new <- factor(modData1$stage_new, levels = c("1", "2", "3_plus"))
class(modData1$stage_new)

stage_dummies <- model.matrix(~ stage_new, data = modData1)[, -1]
stage_dummies


predictors <- modData1[, c("age_at_diagnosis", "gender_1", "p_Tumor", "p_Macrophage", "p_Cytotoxic_T", 
                           "p_T_Helper", "p_B_Cell", "k_Tumor", "k_Cytotoxic_T", "k_B_Cell", "k_Macrophage", 
                           "k_T_Helper", "k_Tumor_CytotoxicT", "k_Tumor_BCell", "k_Tumor_Macrophage", 
                           "k_Tumor_THelper", "k_BCell_CytotoxicT", "k_BCell_Macrophage", "k_BCell_THelper", 
                           "k_CytotoxicT_Macrophage", "k_CytotoxicT_THelper", "k_Macrophage_THelper")]

# Combine predictors and dummy variables
modData2 <- cbind(predictors, stage_dummies)
names(modData2)
```



Kaplan Meier Curves for Lung Cancer Patients
```{r data}

range(filtered_modData$age_at_diagnosis)

hist(filtered_modData$age_at_diagnosis)

filtered_modData_surv <- filtered_modData %>%
  mutate(age_group = case_when(
    age_at_diagnosis < 60 ~ "<60",
    age_at_diagnosis >= 60 & age_at_diagnosis < 70 ~ "60-69",
    age_at_diagnosis >= 70 & age_at_diagnosis < 80 ~ "70-79",
   age_at_diagnosis >= 80 ~ "80+"
   ))



lab_names<- c("<60", "60-69","70-79", "80+")

#Kaplan Meier Curve
km_fit <- survfit(
  Surv(survival_days, survival_status) ~ age_group, data=filtered_modData_surv)

library(RColorBrewer)
display.brewer.all()

kap_lung<-ggsurvplot(km_fit, 
           size=1.5,
           palette = "RdGy",  # Use "Blues" palette
           title = "       Lung Cancer: Patient Survival by Age Group", 
           risk.table =F,
           #legend.labs = lab_names,
           xlab="Time (days)",
           legend.title="Age Group",
           font.x = "bold",
           font.y="bold",
           font.title="bold",
           pval=T)
           #conf.int = TRUE)

kap_lung


barplot(table(filtered_modData_surv$age_group),
        main = "Frequency of Categories",
        xlab = "Category",
        ylab = "Frequency",
        col = "skyblue",
        border = "black")


# Extract the ggplot object from ggsurvplot
kap_lung_plot <- kap_lung$plot

# Save the plot using ggsave with the extracted ggplot object
ggsave("kap_lung_survival_plot.png", plot = kap_lung_plot, width = 7, height = 6)



?cut_number

median=median(filtered_modData_surv$age_at_diagnosis)

filtered_modData_surv <- filtered_modData %>%
  mutate(age_group = case_when(
    age_at_diagnosis < median ~ "group1",
    age_at_diagnosis >= median ~ "group 2"
   ))

filtered_modData_surv$age_group
#Kaplan Meier Curve
km_fit <- survfit(
  Surv(survival_days, survival_status) ~ age_group, data=filtered_modData_surv)


kap_lung<-ggsurvplot(km_fit, 
           size=1.5,
           palette = "viridis",  # Use "Blues" palette
           title = "       Lung Cancer: Patient Survival by Age Group", 
           risk.table =F,
           #legend.labs = lab_names,
           xlab="Time (days)",
           legend.title="Age Group",
           font.x = "bold",
           font.y="bold",
           font.title="bold",
           pval=T)
           #conf.int = TRUE)

kap_lung


```


Cox Model with Elastic Net for Lung Data

Selected alpha and lambda values using cross validation

alpha= 0.05
lambda= 0.6727818
```{r pressure, echo=FALSE}
names(modData1)
#define x values
lung_k <- modData2 %>%
  select(
    age_at_diagnosis,
    stage_new2,
    stage_new3_plus,
    gender_1,
    p_Tumor,
        p_Macrophage,
        p_Cytotoxic_T,
        p_T_Helper,
        p_B_Cell,
         k_Tumor,   
         k_Cytotoxic_T,    
        k_B_Cell,
         k_Macrophage,
         k_T_Helper,
         k_Tumor_CytotoxicT,
         k_Tumor_BCell, 
         k_Tumor_Macrophage,
         k_Tumor_THelper,
         k_BCell_CytotoxicT,
         k_BCell_Macrophage,
         k_BCell_THelper,
         k_CytotoxicT_Macrophage,
         k_CytotoxicT_THelper,
         k_Macrophage_THelper) %>%
  filter(complete.cases(.))


lung_k
#takes it down to a sample size of 146/153


#remove rows from total data set
filtered_modData <- modData[complete.cases(modData %>%
                                             select(p_Tumor,
        p_Macrophage,
        p_Cytotoxic_T,
        p_T_Helper,
        p_B_Cell,
         k_Tumor,   
         k_Cytotoxic_T,    
        k_B_Cell,
         k_Macrophage,
         k_T_Helper,
         k_Tumor_CytotoxicT,
         k_Tumor_BCell, 
         k_Tumor_Macrophage,
         k_Tumor_THelper,
         k_BCell_CytotoxicT,
         k_BCell_Macrophage,
         k_BCell_THelper,
         k_CytotoxicT_Macrophage,
         k_CytotoxicT_THelper,
         k_Macrophage_THelper)), ] 


#fit x and y matricies
x <- as.matrix(lung_k)
x

y <- filtered_modData %>%
  mutate(time=survival_days,
         status=survival_status) %>%
  select(time, status) %>%
  as.matrix()
y

#-------------------------

#choosing alpha for cox model
cvm <- NULL
alpha <- seq(0.05, 1, by = 0.05)
lambda <- list()
lambda_alphamin <- list()
alpha.min <- NULL


set.seed(1)
for(j in 1:50){
  foldid <- sample(1:10, size = nrow(y), replace = T)
  for(i in 1:19){
    a <- alpha[i]
    enetMod.cv <- cv.glmnet(
      x = x, y = y,
      #weights = train$WTSA2YR, 
      #foldid = foldid, 
      alpha = a, 
      #nfolds=10,
      foldid=foldid, 
      #added in line to make it cox
      family="cox",
      #added type.measure
      type.measure = "deviance", 
      #changed standardize from F to T
      standardize = TRUE
      #penalty.factor = c(rep(0, times = 10), rep(1, times = 29))
    )
    cvm[i] <- min(enetMod.cv$cvm)
    lambda[[i]] <- c(enetMod.cv$lambda.min, enetMod.cv$lambda.1se)
  }
  alpha.min[j] <- alpha[which.min(cvm)]
  lambda_alphamin[[j]] <- lambda[[which.min(cvm)]]
  
  print(j)
}

hist(alpha.min, breaks = seq(0, 1, by = 0.05))
summary(alpha.min)
table(alpha.min)


### building cox model using chosen alpha

# if we want to fix demographic variables and only select spatial features
# set age and gender equal to 0 so that model does not penalize them
#penalty_vars <- rep(1, ncol(x))
#penalty_vars
#penalty_vars[c(1, 2, 3)] <- 0


#fit model
cox_lung_fit <- glmnet(x, y, alpha=0.05, family = "cox", standardize = TRUE)

#plot(fit)
plot(cox_lung_fit)


#selecting lambda value using cross validation
set.seed(1)

cvfit_lung_cox <- cv.glmnet(x, y, family = "cox", alpha=0.05, type.measure = "deviance", standardize = TRUE, nfolds=10)

#plot cross validation
plot(cvfit_lung_cox)


# Extract lambda values
lambda_min <- cvfit_lung_cox$lambda.min
lambda_min

set.seed(1)
final_fit<-glmnet(x, y, alpha=0.05, lambda=lambda_min, family = "cox", standardize = TRUE)

final<-coef(final_fit, standardize = TRUE)


?glmnet

pred = predict(final_fit, x, s = lambda_min)
Cindex(pred, y, weights = rep(1, nrow(y)))


#----------------






```




```{r}

k_BCell_CytotoxicT_quantile_25 <- quantile(filtered_modData$k_BCell_CytotoxicT, probs = 0.25)
k_BCell_CytotoxicT_quantile_50 <- quantile(filtered_modData$k_BCell_CytotoxicT, probs = 0.50)
k_BCell_CytotoxicT_quantile_75 <- quantile(filtered_modData$k_BCell_CytotoxicT, probs = 0.75)
k_BCell_CytotoxicT_quantile_25
k_BCell_CytotoxicT_quantile_50



names(filtered_modData)
    filtered_modData   

newData <- with(filtered_modData, data.frame(k_BCell_CytotoxicT = c(k_BCell_CytotoxicT_quantile_25,k_BCell_CytotoxicT_quantile_75),
                                                
  age_at_diagnosis = mean(age_at_diagnosis, na.rm = TRUE),
  stage_new2 = 0,
  stage_new3_plus = 0,
  gender_1 = 0,
  p_Tumor = mean(p_Tumor, na.rm = TRUE),
  p_Macrophage = mean(p_Macrophage, na.rm = TRUE),
  p_Cytotoxic_T = mean(p_Cytotoxic_T, na.rm = TRUE),
  p_T_Helper = mean(p_T_Helper, na.rm = TRUE),
  p_B_Cell = mean(p_B_Cell, na.rm = TRUE),
  k_B_Cell = mean(k_B_Cell, na.rm = TRUE),
  k_Macrophage = mean(k_Macrophage, na.rm = TRUE),
  k_Cytotoxic_T = mean(k_Cytotoxic_T, na.rm = TRUE),
  k_T_Helper = mean(k_T_Helper, na.rm = TRUE),
  k_Tumor_CytotoxicT = mean(k_Tumor_CytotoxicT, na.rm = TRUE),
  k_Tumor_BCell = mean(k_Tumor_BCell, na.rm = TRUE),
  k_Tumor_Macrophage = mean(k_Tumor_Macrophage, na.rm = TRUE),
  k_Tumor_THelper = mean(k_Tumor_THelper, na.rm = TRUE),
  k_BCell_CytotoxicT = mean(k_BCell_CytotoxicT, na.rm = TRUE),
  k_BCell_Macrophage = mean(k_BCell_Macrophage, na.rm = TRUE),
  k_BCell_THelper = mean(k_BCell_THelper, na.rm = TRUE),
  k_CytotoxicT_Macrophage = mean(k_CytotoxicT_Macrophage, na.rm = TRUE),
  k_CytotoxicT_THelper = mean(k_CytotoxicT_THelper, na.rm = TRUE),
  k_Macrophage_THelper = mean(k_Macrophage_THelper, na.rm = TRUE)
))

newData<-as.matrix(newData)
newData

fit <- predict(final_fit, newx = newData, type = "response")

fit

library(glmnet)
library(survival)
library(survminer)

# Create survival plot
ggsurvplot(final_fit)


# , data = newData, 
#            surv.median.line = "hv", # Display median survival time line
#            risk.table = TRUE,      # Display risk table
#            pval = TRUE,            # Show p-value of survival comparison
#            conf.int = TRUE,        # Show confidence intervals
#            ggtheme = theme_minimal())



# Create a survfit object using predicted survival probabilities
surv_obj <- Surv(time = filtered_modData$time, event = lung$status)

names(filtered_modData)

survfit_obj <- survfit(surv_obj ~ predicted_survival_probs)

# Plot using ggsurvplot
ggsurvplot(survfit_obj, conf.int = TRUE, legend.labs = c("Sex=1", "Sex=2"),
           ggtheme = theme_minimal())





# # Survival Curve plot by tumor cell clustering
# #median_k_BCell_CytotoxicT<- median(filtered_modData$k_BCell_CytotoxicT)
# 
# filtered_modData_ovarian_2_median_tumor <- newData_ov %>%
#   mutate(tumor_group = case_when(
#     k_BCell_CytotoxicT< k_BCell_CytotoxicT_quantile_25 ~ "Below 25th Quantile",
#     k_BCell_CytotoxicT >= k_BCell_CytotoxicT_quantile_25 &  k_BCell_CytotoxicT < k_BCell_CytotoxicT_quantile_50 ~ "Between 25th and 50th Quantiles",
#     k_BCell_CytotoxicT>= k_BCell_CytotoxicT_quantile_50 &  k_BCell_CytotoxicT < k_BCell_CytotoxicT_quantile_75 ~ "Between 50th and 75th Quantiles",
#     k_BCell_CytotoxicT > k_BCell_CytotoxicT_quantile_75 ~ "Above 75th"
#   ))




# x_group<- model.matrix(surv_obj ~ tumor_group, filtered_modData_ovarian_2_median_tumor)
# x_group
# 
# final_fit<-glmnet(x_group, surv_obj, alpha=0.95, lambda=lambda_min, family = "cox", standardize = TRUE)
# 
# #final_fit<-glmnet(x_ov_CK_group, surv_obj, alpha=0.05, lambda=lambda_min, family = "cox", standardize = TRUE)
# 
# 
# log_hr <- predict(final_fit, s = lambda_min, type = "link", newx = x_group)
# cum_hazard <- exp(log_hr)
# surv_prob <- exp(-cum_hazard)
# surv_data <- data.frame(
#   time = y_new$time,
#   status = y_new$status,
#   group = filtered_modData_ovarian_2_median_tumor$tumor_group,
#   surv_prob = surv_prob
# )
# 
# surv_data$group <- factor(surv_data$group)
# 
# surv_fit <- survfit(Surv(time, status) ~ group, data = surv_data)
# 
# plot(surv_fit, col = c("blue", "red", "green", "grey"), lty = 1:4, 
#      xlab = "Time", ylab = "Survival Probability", 
#      main = "Survival Curves by Lung Cancer Cell Clustering")
# legend("topright", legend = levels(surv_data$group), col = c("blue", "red", "green", "grey"), lty = 1:3, cex = 0.8)
# 
# 
# 
# 
# 
# 
# 
# 
# 
# # Extra code
# surv_prob <- exp(-pred)
# surv_data <- data.frame(time = y_new$time, surv_prob = surv_prob)
# 
# surv_fit <- survfit(Surv(y_new$time, y_new$status) ~ 1, data = surv_data)
# 
# plot(surv_fit, xlab = "Time", ylab = "Survival Probability", main = "Survival Curves")



```


Check for multicollinearity
```{r corr}
library(corrplot)

corr<-cor(lung_k, method="pearson")

corrplot(corr, method="number", type="upper", )

```



AFT model for Lung Data

alpha = 0.05
lambda = 
```{r aft}
#mutate to log time
ydf <- as.data.frame(y)
logY <- ydf %>%
  mutate(time = log(time),
         status = status) 

startT <- Sys.time()
#select alpha using cross validation
cvm <- NULL
alpha <- seq(0.05, 1, by = 0.05)
lambda <- list()
lambda_alphamin <- list()
alpha.min <- NULL

set.seed(1)
for(j in 1:300){
  folds <- sample(1:10, nrow(y), replace = T)
  foldid <- lapply(1:10, function(x){which(folds == x)})
  for(i in 1:19){
    a <- alpha[i]
    enetMod.cv <- penAFT.cv(x_mat, logY$time, logY$status, 
                           nlambda = 50,
                           penalty = NULL,
                           alpha=a,
                           cv.index=foldid,
                           standardize=T
    )
    cvm[i] <- min(enetMod.cv$cv.err.linPred)
    lambda[i] <- enetMod.cv$lambda.min
  }
  alpha.min_aft[j] <- alpha[which.min(cvm)]
  lambda_alphamin[j] <- lambda[which.min(cvm)]
  
  print(j)
}
Sys.time() - startT



hist(alpha.min_aft, breaks = seq(0, 1, by = 0.05))
summary(alpha.min_aft)
alpha_freq_table_aft<-table(alpha.min_aft)

alpha_freq_table_aft

chosen_alpha<-which.max(alpha_freq_table_aft)
alpha_with_max_freq<-names(alpha_freq_table_aft)[chosen_alpha]
alpha_with_max_freq




aft_fit_plot <- penAFT(x, logY$time, logY$status, 
                           penalty = NULL,
                       nlambda=50,
                           alpha=0.05,
                           lambda = NULL,
                           standardize=T)
penAFT.trace(aft_fit_plot)

#chose 0.05

set.seed(1)
aft_mod_pen_aft<-penAFT.cv(x, logY$time, logY$status, 
                           nlambda = 50,
                           #penalty = EN,
                           alpha=0.05,
                           nfolds=10, 
                           standardize=T)



#refit model with lambda min
aft_best_lambda<-aft_mod_pen_aft$lambda.min
aft_best_lambda




?penAFT

# fit model using chosen alpha 
aft_fit_final <- penAFT(x, logY$time, logY$status, 
                           penalty = NULL,
                           alpha=0.05,
                           lambda = aft_best_lambda,
                           standardize=T)




coeff<-penAFT.coef(aft_fit_final, lambda = aft_best_lambda)
coeff

penAFT.plot(aft_mod_pen_aft)

names<-c("age_at_diagnosis",
"gender_1",
"stage2",
"stage3",
"p_Tumor",
"p_Macrophage",
"p_Cytotoxic_T",
"p_T_Helper",
"p_B_Cell",
"k_Tumor",   
"k_Cytotoxic_T",    
"k_B_Cell",
"k_Macrophage",
"k_T_Helper",
"k_Tumor_CytotoxicT",
"k_Tumor_BCell", 
"k_Tumor_Macrophage",
"k_Tumor_THelper",
"k_BCell_CytotoxicT",
"k_BCell_Macrophage",
"k_BCell_THelper",
"k_CytotoxicT_Macrophage",
"k_CytotoxicT_THelper",
"k_Macrophage_THelper")

results <- data.frame(Variable = names, Coefficient = coeff)
results





coef_best_lambda <- penAFT.coef(aft_mod_pen_aft, lambda = "lambda.min")

?penAFT.predict


pred_aft = penAFT.predict(aft_fit_final, x, lambda = aft_best_lambda)
Cindex(pred_aft, y, weights = rep(1, nrow(y)))







```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
